\Title

### 2013-06-06 synmut.tex
Text changes + fig 3A for color blind

\begin{abstract}
\noindent
Intrapatient HIV-1 evolution is driven by the adaptive immune system
resulting in rapid evolution of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
or neutralizing antibodies target a new epitope, the virus often escapes via
nonsynonymous mutations that impair recognition. Synonymous mutations do not
affect this interplay and are often assumed to be neutral. We test this
assumption by tracking synonymous mutations in 
longitudinal intrapatient data from the \shankaregion{} part of the envelope
gene (\env{}).  We find that most synonymous mutations are lost even though
they often reach high frequencies in the viral population suggesting a
cost to the virus. Using published data
from SHAPE assays, we find that synonymous mutations that disrupt base pairs
in RNA stems flanking the variable loops of gp120 are more likely to be lost than other
synonymous changes, suggesting a function of these RNA hairpins in HIV-1.
Computational modeling indicates that these synonymous mutations have a
selection coefficient of the order of $-0.002$ and that they are rise to
high frequency by genetic hitchhiking on neighboring beneficial
variants. This weak selection against synonymous substitutions does not
result in a strong pattern of conservation in cross-sectional data, but
slows down the rate of evolution considerably. Our findings are consistent with the
notion that large scale patterns of RNA structure are functionally
relevant, while the precise base pairing pattern is not.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by the high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}.  Escape mutations
in epitopes targeted by CTLs are typically observed during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120, which change to avoid recognition by
nAbs. Escape mutations in \env, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

Escape mutations are selected because they change the amino acid sequence of
viral proteins in a way that reduces antibody binding or epitope presentation.
Conversely, synonymous mutations do not modify the viral proteins and are
commonly used as approximately neutral markers in studies of viral evolution.
Neutral markers are very useful as a negative control for detecting selected
sites \citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
maintaining protein function and avoiding the adaptive immune recognition,
however, the HIV-1 genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate
at the RNA level and are sensitive to synonymous changes. For example, the HIV-1
\rev{} response element (RRE) in \env{} enhances nuclear export of full length
or partially spliced viral transcripts via a complex hairpin RNA structure
\citep{fernandes_hiv-1_2012}. Another well studied case is the interaction
between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and is bound by a pseudoknotted RNA structure in the viral 5' untranslated
region~\citep{barat_interaction_1991, paillart_vitro_2002}. 
Besides these well characterized and conserved structures, large parts of the virus
genome are folded into structures of unknown function
\citep{watts_architecture_2009}. These poorly characterized RNA
structures are conserved to varying degree between HIV-1 and
SIV in the sense that corresponding regions tend to be part of
of similar structural elements. Individual base pairings, however, are
not conserved \citep{pollom_comparison_2013}.

@@

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that, in the region of the V loops, a large fraction of these mutations
is deleterious. Despite their fitness cost, deleterious synonymous mutations
rise in frequency in the viral population via genetic hitchhiking due to limited
recombination in HIV-1 populations~\citep{neher_recombination_2010,
batorsky_estimate_2011}. We show a strong correlation between the fate of a
synonymous mutation and the surrounding RNA structure. We then compare our
observations to computational models and obtain estimates for the effect of
synonymous mutations on viral fitness. Extending the analysis of fixation
probabilities to the nonsynonymous mutations, we show that time-dependent
selection or strong competition of escape mutations inside the same epitope are
necessary to explain the observed patterns of fixation and loss.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Due to the large population size and the high mutation rate, every
possible single nucleotide variant (SNV) is produced multiple time per
day \citep{coffin_hiv_1995}. Some of these SNVs rise to high enough
frequency that they are observed in a sample of sequences. SNVs rise or fall in 
frequency because of (i) their own effect on fitness and escape and (ii)
their association to genetic backgrounds and (iii) stochastic fluctuations
(genetic drift). We study the dynamics of sample frequencies, $\nu$, of SNVs
(SNV), i.e., the fraction $\nu$ of the sequences in a sample carrying
the SNV. When an SNVs is present in all sequences 
at a certain time point, we say it has ``fixed''; when it is completely absent,
we say it was ``lost'' or is ``extinct''.

Most positions are only transiently variable and SNVs will either fix or
will be lost -- at least in small samples. Given a SNV is at a certain
frequency $\nu$, the probability of fixation is higher for beneficial
SNVs than for neutral ones, which in turn fix more frequently than
deleterious ones. Since a neutral mutation, by definition, has no effect
on the fate of the viruses carrying it, the fixation probability of a
neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
$\pfix(\nu) = \nu$; it goes extinct with probability $1-\nu$. For
instance, if a neutral SNV is observed in half of the sequences, it will
fix with a probability of 50\% (see inset in figure
\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
independent of most model assumptions and is only affected if neutral
SNVs are associated preferentially with either high or low fitness
virus.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Many synonymous SNVs in \shankaregion{} are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We studied the dynamics and fate of synonymous SNVs more quantitatively by
analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
\citet{liu_selection_2006} as well as three patients from
\citet{bunnik_autologous_2008} (patients with strong viral population structure
were excluded from the analysis; see methods and \figurename~S\PCApat). The
former data set is restricted to the \shankaregion{} region of \env, while the
data from \citet{bunnik_autologous_2008} cover most of \env.  We considered all
SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
$t$, and calculated the fraction that are still observed at later times $t+\Delta
t$. Plotting this fraction against the time interval $\Delta t$, we see that
most synonymous SNVs segregate for roughly one year and are lost much more
frequently than expected under neutrality (panel \ref{fig:fixp1}). The long-time
probability of fixation, $\pfix$, is shown as a function of the initial
frequency $\nu_0$ in panel \ref{fig:fixp2}. We found that $\pfix$ of synonymous
variants is far below the neutral expectation in \shankaregion (red line).
Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
we found no such reduction in $\pfix$ (green line). Restricted to the
\shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
are fully compatible with data from \citet{shankarappa_consistent_1999}. The
nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
line) -- a point to which we will come back below.

@@

When interpreting these results for the fixation probabilities, it is important
to note that we focused on SNVs that have already reached high frequencies. In
HIV-1 infection, most SNVs remain very rare throughout; they are not considered
here. Synonymous SNVs can reach high frequencies either through genetic
drift or genetic hitchhiking on escape variants (see below); very deleterious
variants will never reach high frequencies in the first place. Hence, our
analysis indicates that, among all synonymous SNVs that somehow reach high
frequencies, most of those in \shankaregion{} are deleterious, while
those in the rest of \env{} tend to be neutral.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Synonymous mutations in \shankaregion{} tend to disrupt RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for lack of fixation of synonymous SNVs in
\shankaregion{} is secondary structures in the viral RNA, the disruption of which
is deleterious to the virus \citep{forsdyke_reciprocal_1995,
snoeck_mapping_2011, sanjuan_interplay_2011}.

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases 
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
variable regions form stems. We aligned the within-patient sequence samples to
the reference NL4-3 strain used in \citet{watts_architecture_2009} and assigned
SHAPE reactivities to most positions in the alignment. We then calculated the
distributions of SHAPE reactivities for synonymous SNVs that fixed or were
subsequently lost (only SNVs with frequencies above 15\%). As shown in
\FIG{SHAPEA}, the reactivities of fixed SNVs (red histogram) are systematically
larger than those of lost SNVs (blue) (Kolmogorov-Smirnov test on the cumulative
distribution, $p\approx 0.002$). In other words, SNVs that are likely to
break RNA helices are also more likely to revert and finally be lost from the
population, restoring the helix. Note that this analysis will be sensitive only
at positions where the base pairing pattern of NL4-3 agrees with that of each
patient's initial consensus sequence -- it is thus statistically conservative.
As a control, we also calculate the distribution of SHAPE reactivities for SNVs
that never reach high frequencies (green). This set is a mixture of neutral and
deleterious SNVs and, as expected, their distribution lies between those of
fixed and lost high-frequency SNVs.

To test the hypothesis that synonymous SNVs in \shankaregion{} are lost because they
break stems in the conserved stretches between the V loops, we considered
separately SNVs in V loops and conserved flanks. The greatest
depression in fixation probability is observed in the conserved stems, while the
V loops show little deviation from the neutral signature, see
\FIG{SHAPEB}. This is suggestive of important RNA helices in conserved
regions between the V loops.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Deleterious SNVs reach high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

While the observation that some fraction of synonymous SNVs is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome (in \shankaregion{} but not in the rest of \env{}; compare the red
triangle line versus the green square line in \FIG{fixp2}).
The region \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious SNVs
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is apparent in
\FIG{aft}, which shows that many SNVs change rapidly in frequency as a
flock.

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the
parameters, see \FIG{simsfig}. Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Specifically, \FIG{simsfig} shows
parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
P_\text{interm} < 0.010$. This subset of parameters indicates that a high fraction
($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
0.002$, consistent with the fixation/extinction times estimated above
(see \FIG{fixp1}).

The bounds on the fraction of synonymous SNVs that are deleterious and
their effect sizes are plausible:
(i) A substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
the majority of SNVs reaching high frequency are neutral and no depression is
observed; (ii) in order to hitchhike, the deleterious effect size has to be much
smaller than the escape rate, otherwise the double mutant (with both the escape
mutation and the deleterious synonymous one) has little or no fitness advantage
over the wildtype virus. Consistent with this argument, larger deleterious
effects in \FIG{simsfig} correspond to larger escape rates; and (iii) SNVs
with a deleterious effect smaller than approximately $0.001$ behave neutrally,
consistent with the typical coalescent times observed in HIV-1.

The above simulations show that hitchhiking with favored nonsynonymous
variants can explain the observation of
deleterious synonymous SNVs that rarely fix. However, in a simple model where
nonsynonymous escape mutations are unconditionally beneficial, they almost
always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the blue line in \FIG{fixp2}: in an HIV-1
infection, nonsynonymous SNVs at high frequency often disappear again, even
though many are at least transiently beneficial. Inspecting the trajectories of
nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
possible mechanisms that could explain the transient rise of
nonsynonymous SNVs are time-dependent selection and within-epitope competition.

If the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In
support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
antibody responses against escape mutants have been reported. These responses are
delayed by a few months, roughly matching the average time needed by an escape
mutant to rise from low to high frequency. To model this type of behavior, we
assumed that antibody responses against escape SNVs arise with a rate
proportional to the frequency of the escape SNV and abolish the benefit of the
escape mutations. As expected, this type of time-dependent selection retained the
potential for hitchhiking, but reduced fixation of nonsynonymous SNVs.
\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is not
affected by this change, while $\pfix$ of nonsynonymous SNVs approaches the
diagonal as the rate of recognition of escape mutants is increased.

Several different escape mutations that arise within the 
same epitope have similar effects. Their benefits
are not additive, because each of them is essentially sufficient to escape and
no additional benefit is gained from combining them. As a consequence, several
escape SNVs rise to high frequency rapidly, while the one with the smallest cost
in terms of replication, packaging, etc.~is most likely to eventually fix, while
all others are lost. The emergence of multiple competing escape SNVs in HIV-1
infections has been shown \citep{moore_limited_2009, bar_early_2012}. 
Similarly, this scenario has been explicitly observed in the evolution of
resistance to 3TC, where the mutation M184V is often preceded by M184I
\citep{hedskog_dynamics_2010}. We implemented within epitope competition 
in the model by allowing multiple escape mutations per epitope that do
not provide additional benefit to the virus when combined. Again, we found that the potential for
hitchhiking is retained by within epitope competition but that the
fixation probability of nonsynonymous SNVs is reduced. With roughly six
mutations per epitope, the simulation data are compatible with observations; see
\figurename~S\withinepi. The two scenarios, time-dependent selection and
competition between equivalent escape pathways, are not exclusive and possibly
both important in HIV-1 evolution.

>>>>

\begin{abstract}
\noindent
Intrapatient HIV-1 evolution is driven by the adaptive immune system
resulting in rapid adaptation of HIV-1 proteins. When cytotoxic CD8${}^+$ T-cells
or neutralizing antibodies target a new epitope, the virus often escapes via
nonsynonymous mutations that impair recognition. Synonymous mutations do not
affect this interplay and are often assumed to be neutral. We test this
assumption by tracking the abundance of synonymous mutations in 
longitudinal intrapatient data from the \shankaregion{} part of the \env{}
gene. We find that most synonymous variants are lost even though
they often reach high frequencies in the viral population, suggesting a
cost to the virus. Using published data
from SHAPE assays, we find that synonymous mutations that disrupt base pairs
in RNA stems flanking the variable loops of gp120 are more likely to be lost than other
synonymous changes; these RNA hairpins might be important for HIV-1.
Computational modeling indicates that (i) most synonymous mutations in this
genomic region have a selection coefficient of the order of $-0.002$, and
(ii) synonymous variants rise to high frequency by genetic hitchhiking on neighboring beneficial
variants. This weak selection against synonymous substitutions does not
result in a strong pattern of conservation in cross sectional data, but
slows down the rate of evolution considerably. Our findings are consistent with the
notion that large scale patterns of RNA structure are functionally
relevant, while the precise base pairing pattern is not.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by HIV-1's high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. Escape mutations
in epitopes targeted by CTLs are typically observed during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120 (V loops), which change to avoid recognition by
nAbs. Escape mutations in \env{}, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes autologous virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

Escape mutations are selected because they change the amino acid sequence of
viral proteins in a way that reduces antibody binding or epitope presentation.
Conversely, synonymous mutations do not modify the viral proteins and are
commonly used as approximately neutral markers in studies of viral evolution,
i.e. as a negative control for detecting selected
sites \citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
maintaining protein function and avoiding the adaptive immune recognition,
however, the HIV-1 genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate
at the RNA level and are sensitive to synonymous changes. For example, the HIV-1
\rev{} response element (RRE) in \env{} enhances nuclear export of full length
or partially spliced viral transcripts via a complex hairpin RNA structure
\citep{fernandes_hiv-1_2012}. Another well studied case is the interaction
between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and is bound by a pseudoknotted RNA structure in the viral 5' untranslated
region~\citep{barat_interaction_1991, paillart_vitro_2002}. 
Besides these well characterized and conserved structures, large parts of the virus
genome are folded into structures of unknown function
\citep{watts_architecture_2009}. These poorly characterized RNA
structures are conserved to varying degree between HIV-1 and
SIV: corresponding regions tend to be part of
of similar structural elements, but individual base pairings differ widely \citep{pollom_comparison_2013}.

@@

In this paper, we characterize the dynamics of synonymous mutations in \env{}
and show that, in the region of the V loops, a large fraction of these mutations
is deleterious. Despite their fitness cost, deleterious synonymous variants
rise in frequency in the viral population via genetic hitchhiking due to limited
recombination in HIV-1 populations~\citep{neher_recombination_2010,
batorsky_estimate_2011}. We show a strong correlation between the fate of a
synonymous variant and the surrounding RNA structure. We then compare our
observations to computational models and obtain estimates for the effect of
synonymous mutations on viral fitness.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DO WE WANT TO KEEP THIS LAST SENTENCE?
Extending the analysis of fixation
probabilities to the nonsynonymous variants, we show that time-dependent
selection or competition of escape mutations inside the same epitope are
necessary to explain the observed patterns of spread and extinction.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Due to the large population size and the high mutation rate, every
possible single nucleotide variant (SNV) is produced multiple times per
day \citep{coffin_hiv_1995}. Some of these variants rise to high enough
frequency that they are observed in a sample of sequences. SNVs rise or fall in 
frequency because of three reasons: (i) their own effect on fitness and escape; (ii)
their association to genetic backgrounds; (iii) stochastic fluctuations
(genetic drift). We study the dynamics of sample frequencies, $\nu$, of SNVs,
i.e., the fraction $\nu$ of the sequences in a sample carrying
the variant. When an SNVs is present in all sequences 
at a certain time point, we say it has ``fixed''; when it is completely absent,
we say it was ``lost'' or is ``extinct''.

Most positions are only transiently variable and variants will either fix or
will be lost -- at least in small samples. Given an SNV is at a certain
frequency $\nu$, the probability of fixation is higher for beneficial
SNVs than for neutral ones; in turn, neutral variants fix more frequently than
deleterious ones. The fixation probability of a
neutral SNV at frequency $\nu$ is the frequency itself, i.e.,
$\pfix(\nu) = \nu$; it goes extinct with probability $1-\nu$. For
instance, if a neutral SNV is observed in half of the sequences, it will
fix with a probability of 50\% (see inset in figure
\ref{fig:aftsyn}). The fixation probability of neutral SNVs is
independent of most model assumptions and is only affected if neutral
SNVs are associated preferentially with either high or low fitness
virus.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Many synonymous SNVs in \shankaregion{} are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We studied the dynamics and fate of synonymous variants more quantitatively by
analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
\citet{liu_selection_2006} as well as three patients from
\citet{bunnik_autologous_2008} (patients with strong viral population structure
were not considered; see methods and \figurename~S\PCApat). The
former data set is restricted to the \shankaregion{} region of \env, while the
data from \citet{bunnik_autologous_2008} cover most of \env.  We considered all
SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
$t$, and calculated the fraction that are still observed at later times $t+\Delta
t$. Plotting this fraction against the time interval $\Delta t$, we see that
most synonymous SNVs segregate for roughly one year and are lost much more
frequently than expected under neutrality (panel \ref{fig:fixp1}). The long-time
probability of fixation, $\pfix$, is shown as a function of the initial
frequency $\nu_0$ in panel \ref{fig:fixp2}. As a control, the neutral result is
shown as a black dashed line. We found that $\pfix$ of synonymous
variants is far below the neutral expectation in \shankaregion{} (red triangle line).
Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
we found no such reduction in $\pfix$ (green square line). Restricted to the
\shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
are fully compatible with data from \citet{shankarappa_consistent_1999}:
fixation is rare. The
nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
circle line) -- a point to which we come back below.

@@

When interpreting these results for the fixation probabilities, it is important
to note that we focused on SNVs that have already reached high frequencies. In
HIV-1 infection, most SNVs remain very rare throughout: they are not considered
here. Synonymous SNVs can reach high frequencies either through genetic
drift or genetic hitchhiking on escape variants (see below); very deleterious
variants will never reach high frequencies in the first place. Hence, our
analysis indicates that, among all synonymous SNVs that somehow reach high
frequencies, most of those in \shankaregion{} are deleterious, while
those in the rest of \env{} tend to be neutral.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Synonymous mutations in \shankaregion{} tend to disrupt RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for a reduced fixation of synonymous variants in
\shankaregion{} is secondary structures in the viral RNA, the disruption of which
is deleterious to the virus \citep{forsdyke_reciprocal_1995,
snoeck_mapping_2011, sanjuan_interplay_2011}.

The propensity of nucleotides in the HIV-1 genome to form base pairs has been
measured using the SHAPE assay, a biochemical reaction preferentially altering
unpaired bases 
\citep{watts_architecture_2009}. The SHAPE assay has shown that the variable
regions V1-V5 tend to be unpaired, while the conserved regions between those
stretches form stems. We aligned the sequences from each patient to
the reference NL4-3 strain used in \citet{watts_architecture_2009} and assigned
SHAPE reactivities to most positions in the alignment. We then calculated the
distributions of SHAPE reactivities for synonymous SNVs that fixed or were
lost (only variants reaching frequencies above 15\%). As shown in
\FIG{SHAPEA}, the reactivities of fixed SNVs (red histogram) are systematically
larger than those of lost SNVs (blue) (Kolmogorov-Smirnov test on the cumulative
distribution, $p\approx 0.002$). In other words, SNVs that are likely to
break RNA helices are also more likely to revert and finally be lost from the
population, restoring the helix. Note that this analysis will be sensitive only
at positions where the base pairing pattern of NL4-3 agrees with that of each
patient's initial consensus sequence -- it is thus statistically conservative.
As a control, we also calculate the distribution of SHAPE reactivities for SNVs
that never reach high frequencies (green). This set is a mixture of neutral and
deleterious SNVs and, as expected, its distribution lies between those of
fixed and lost high-frequency variants.

To test the hypothesis that synonymous variants in \shankaregion{} are lost because they
break stems in the conserved stretches between the V loops, we considered
separately SNVs in V loops and their flanks. The greatest
depression in fixation probability is observed in the conserved stems, while the
V loops show little deviation from the neutral signature, see
\FIG{SHAPEB}. This is suggestive of important RNA helices in conserved
regions between the V loops.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Deleterious SNVs reach high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the observation of some deleterious synonymous variants is not fully
unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome (in \shankaregion{} but not in the rest of \env{}; compare the red
triangle line versus the green square line in \FIG{fixp2}).
The region \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious variants
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is apparent in
\FIG{aft}, which shows that many SNVs change rapidly in frequency as a
flock.

@@

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution with random choices for the
parameters, see \FIG{simsfig}. Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Specifically, \FIG{simsfig} shows
parameter combinations for which we found $A_\text{syn} < -0.15$ and $0.0025 <
P_\text{interm} < 0.010$. This subset of parameters indicates that a high fraction
($\gtrsim 0.8$) of sites has to be deleterious with effect size $|s_d| \sim
0.002$. This is consistent with the fixation/extinction times estimated above
(see \FIG{fixp1}).

The bounds on the fraction of deleterious synonymous SNVs and
on their effect sizes are plausible:
(i) a substantial depression in $\pfix$ requires pervasive deleterious SNVs, otherwise
the majority of SNVs reaching high frequency are neutral and no depression is
observed; (ii) in order to hitchhike, the deleterious effect size has to be much
smaller than the escape rate, otherwise the double mutant (with both the escape
mutation and the deleterious synonymous one) has little or no fitness advantage
over the wildtype virus. In agreement with this argument, larger deleterious
effects in \FIG{simsfig} correspond to larger escape rates; and (iii) SNVs
with a deleterious effect smaller than approximately $0.001$ behave neutrally,
as expected from the typical coalescent times observed in HIV-1.

The above simulations show that hitchhiking with favoured nonsynonymous
variants can explain the observation of
deleterious synonymous SNVs that rarely fix. However, in a simple model where
nonsynonymous escape mutations are unconditionally beneficial, escape mutations almost
always fix once they reach high frequencies, i.e., $A_{\mathrm{nonsyn}}$ is well
above zero. This is incompatible with the blue circle line in \FIG{fixp2}: in an HIV-1
infection, nonsynonymous SNVs at high frequency often disappear again, even
though many are at least transiently beneficial. Inspecting the trajectories of
nonsynonymous SNVs suggests the rapid rise and fall of many SNVs. Two
possible mechanisms that could explain the transient rise of
nonsynonymous SNVs are time-dependent selection and within-epitope competition.

If the immune system
recognizes the escape mutant before its fixation, the mutant might cease to be
beneficial and disappear soon, despite its quick initial rise in frequency. In
support of this idea, in \citet{richman_rapid_2003, bunnik_autologous_2008},
antibody responses against escape mutants have been reported. These responses are
delayed by a few months, roughly matching the average time needed by an escape
mutant to rise from low to high frequency. To model this type of behavior, we
assumed that antibody responses against escape SNVs arise with a rate
proportional to the frequency of the escape variant and abolish the benefit of the
escape mutations. As expected, this type of time-dependent selection retained the
potential for hitchhiking, but reduced fixation of nonsynonymous SNVs.
\figurename~S\timedependence~shows that $\pfix$ of synonymous SNVs is not
affected by this change, while $\pfix$ of nonsynonymous SNVs approaches the
diagonal as the rate of recognition of escape mutants is increased.

Several different escape mutations that arise within the 
same epitope have similar effects. Their benefits
are not additive, because each of them is essentially sufficient to escape and
no additional benefit is gained from combining them. As a consequence, several
escape SNVs rise to high frequency rapidly, while the one with the smallest cost
in terms of replication, packaging, etc.~is most likely to eventually fix, while
all others are lost. The emergence of multiple competing escape variants in HIV-1
infections has been shown \citep{moore_limited_2009, bar_early_2012}. 
Similarly, this scenario has been explicitly observed in the evolution of
resistance to 3TC, where the mutation M184V is often preceded by M184I
\citep{hedskog_dynamics_2010}. We implemented within epitope competition 
in the model by allowing for multiple escape mutations per epitope that do
not provide additional benefit to the virus when combined. Again, we found that the potential for
hitchhiking is retained by within-epitope competition but that the
fixation probability of nonsynonymous SNVs is reduced. With roughly six
mutations per epitope, the simulation data are compatible with observations; see
\figurename~S\withinepi. The two scenarios, time-dependent selection and
competition between equivalent escape pathways, are not exclusive and possibly
both important in HIV-1 evolution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By analyzing the fate of single nucleotide variants (SNVs) in
longitudinal data of HIV-1 \env{} evolution, we demonstrated selection
against synonymous substitutions in the relatively conserved regions
C2-C4 of the \env{} gene. Computational modeling suggested that these
SNVs have deleterious effects on the order of $0.002$ and that they are
brought to high frequency through linkage to adaptive mutations.
Comparison with biochemical studies of base pairing propensity in an RNA
genome of HIV-1 indicated that these mutations are deleterious, at least
in part, because they disrupt RNA secondary structures. Computer models
of RNA folding predict stable hairpins in these regions that
have been suggested to be functional and termed ``insulating
stems'' \citep{watts_architecture_2009, sanjuan_interplay_2011}.
The weak selection against synonymous variants is compatible with the
negative results of \textit{In vitro} replication assays investigating
the fitness effects of small RNA hairpins in HIV-1
\citep{knoepfel_role_2013}: it would take hundreds of cell culture
passages to detect fitness effects of the order of one per mille. The longitudinal data,
however, spans many years, hence our analysis is able to quantify the
subtle fitness effect of RNA structure within single infections.
